\chapter{Motivation}
\textit{Code formatting}, also known as \textit{code beautification} or \textit{prettyprinting},
has long been a staple in software development tooling.
Researchers have noticed its necessity and started working on solutions decades ago,
\autocite{syntaxDirectedPrettyprinting}
and the widespread availability of such coding assistance has increased ever since.

Some languages --- such as Java --- are commonly written in an integrated development environment (IDE).
The emergence of those full-fledged tooling packages has helped render formatting tools ubiquitous.
Other languages --- such as JavaScript --- are more commonly written using a rather lightweight editor,
often because the language is less verbose or simply for historical reasons.
But even some of those editors --- such as the web development-focused Microsoft Visual Studio Code ---
now include formatting tools in their core distributions together with
Git version control integration, syntax highlighting and other tooling that is deemed essential,
disposing the need to install an additional plugin that provides code formatting support.
\autocite{vsCodeJavascriptDoc}

These formatters usually work by lexing and parsing the input code
in a process similar to the one a compiler for the same language would utilize,
but then instead of proceeding to transform or semantically analyze the generated syntax tree,
they use the syntactic information obtained by parsing
to print out the token stream obtained by lexing
in a defined format that humans consider readable and comprehensible,
optionally preserving the shape of the original code to some degree.
\autocite{prettyGoodFormattingPipeline}

We introduce a procedure that skips the parsing step and
build a formatter that can move on from the lexing step straight to printing.
While this procedure will generate less information about the code we are operating on and
thus necessarily limit our capability to generate code
formatted in a consistent and customizable output format,
our formatter will become quite easy to adapt to vastly different languages and
we can potentially achieve higher formatting performance with increased code throughput.

In the following chapter, we will roughly outline how code formatters typically work.
Once we have established fundamental comprehension of their technical background,
we can then move on to propose an alternative method that relinquishes parsing.
We address potential benefits and uses, but also drawbacks of this method,
and afterwards develop its implementation for a concrete language.
